---
title: "UFO Geostatistics"
author: "Michael Ramsey"
date: "February 21, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Statistics of UFO Data 

The National UFO Reporting Center (NUFORC) is an organization in the United States that investigates UFO sightings and/or alien contacts. NUFORC has been in continuous operation since 1974 when it was founded by Robert J. Gribble. It has catalogued almost 90,000 reported UFO sightings over its history, most of which were in the United States. The data analyzed here contains information on 64,506 UFO sightings over the last century. The data contain lon/lat coordinates, state, shape of the UFO, the date and time as well as the duration (in seconds) of the sighting. We will investigate the spatial structure of the dataset. Here is a snapshot of the data.

```{r echo = F, message = F, warning = F}
# Load tidyverse
library(tidyverse)
library(urbnmapr)
library(gridExtra)
library(maps)
library(fields)
library(geoR)
library(mapproj)
library(gridExtra)
library(cowplot)

# Load workspace
load("UFO_scrubbed.RData")
grid.table(head(ufo),rows = NULL)

# Edit the blank cells for shape
ufo$shape <- sub("^$", "Blank", ufo$shape)
```

# Data Visualization

#### Sightings by state

We start by displaying a map of the United States where each point represents the geographic location of a ufo sighting. The second U.S. map displays a heat-map by state, where the color of the state represents the number of sightings in that state. Note that this map is slightly misleading, since we large states will naturally have more sightings than smaller ones. I should normalize by state population, however I did not load the data. It seems that the sigtings are concentrated around areas of large population. We points for all of the cities with population over 40,000 in red.

```{r echo = F, message = F, warning = F}

# Count sightings by state
ufo_state <- ufo %>%
  group_by(state) %>%
  summarise(Count = n()) %>%
  rename(state_abbv = state) %>%
  mutate(state_abbv = toupper(state_abbv)) 

# Join the data
urbn <- urbnmapr::states %>%
  left_join(ufo_state, by = 'state_abbv')

# Get city data
cities <- us.cities %>%
  filter(long > -130)

# Map US data with heat of sightings
ggplot() + 
  geom_polygon(data = urbn, 
               aes(x = long, y = lat, group = group),
		           color = 'grey10', fill = 'white') +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  geom_point(data = ufo, aes(x = longitude, y = latitude), size = .05) +
  geom_point(data = cities, aes(x = long, y = lat), color = 'red', size = .5) +
  ylab('Latitude') +
  xlab('Longitude')

# Map US data with heat of sightings
ggplot() + 
  geom_polygon(data = urbn, 
               aes(x = long, y = lat, group = group, fill = Count),
		           color = 'black') +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  ylab('Latitude') +
  xlab('Longitude') +
  scale_fill_continuous('Sightings') +
  scale_color_gradient2(low="blue", mid="white",
                     high="red", space ="Lab" )
```

The location of 

#### Counts of sightings by shape

```{r echo = F, message = F, warning = F}

# Bar plots of shape
ggplot(data = ufo, aes(x = shape)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  ggtitle('Counts of UFOs by shape')
```

We see that, by far, the most common shape is "light", followed by "triangle", then "circle". There are 29 different shapes recorded.

#### Distriubtion of sighting lengths

Sighting legth was converted to log scale due to the large skewness in the data.

```{r echo = F, message = F, warning = F}

# Remove log-durations that are too large or too small
ufo <- ufo %>%  
  filter(duration.seconds > 1) %>%
  filter(log(duration.seconds) <= 10) %>%
  mutate(logdur = log(duration.seconds))

# Histogram of duration
ggplot(data = ufo, aes(x = log(duration.seconds))) + 
  geom_histogram(bins = 15) + 
  scale_x_continuous(limits = c(0,8)) +
  ggtitle('Histogram of log-duration')
```

I also provide a summary of the duration and log-duration for reference on these values.

```{r echo = F, message = F, warning = F}

# Print summary of duration and log-duartion
sum1 <- summary(ufo$duration.seconds)
sum_names <- names(sum1)
sum1 <- as.vector(sum1)
sum2 <- as.vector(summary(log(ufo$duration.seconds)))
sum_data <- data.frame(Statistic = sum_names,
                       Duration = unname(sum1),
                       Logduration = unname(sum2))
grid.table(sum_data, rows = NULL)
```

# Spatial Modeling

#### Spatial dependence of duration

We start by constructing a heat map for the states.

```{r echo = F, message = F, warning = F}

# LIST OF STATES TO PLOT
state_list = c('co','mn','ny','wa','ma','fl')

# Extract sightings for states
state_data <- ufo %>%
  filter(state %in% state_list)

# Get maps for state
state_maps <- urbn %>%
  filter(state_abbv %in% toupper(state_list))

# Map US data with heat of sightings
state_plots <- list()
for (i in c(1:length(state_list))) 
{
  # Get particular state map
  temp_map <- state_maps %>%
    filter(state_abbv == toupper(state_list[i]))
  
  # Get state sigtings data
  temp_data <- state_data %>%
  filter(state == state_list[i])
  
  # Plot the state
  state_plots[[i]] <- ggplot() + 
  geom_polygon(data = temp_map, 
               aes(x = long, y = lat, group = group),
		           color = 'grey10', fill = 'white') +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  geom_point(data = temp_data, aes(x = longitude, y = latitude,
                            color = logdur), size = 1)+
  scale_color_gradientn(name = 'log-duration',
                        colours = c("darkblue","lightblue","green","yellow","red")) +
  guides(color = F) +
  labs(x=NULL, y=NULL) +
  theme(plot.margin=unit(c(0,0,0,0),"mm"))
}

# Plot the states
grid.arrange(state_plots[[1]], state_plots[[2]],
             state_plots[[3]], state_plots[[4]],
             state_plots[[5]], state_plots[[6]], ncol = 2)

```

#### Fields package

Using the fields package, we get the following plot.

```{r echo = F, message = F, warning = F}

# Map US data with heat of sightings
par(mfrow=c(3,2),
    mar = c(2,2,1,1),
    xpd = NA)

for (i in c(1:length(state_list))) 
{
  # Get state sigtings data
  temp_data <- state_data %>%
  filter(state == state_list[i])
  
  # Plot state with the fields package
  quilt.plot(x = temp_data$longitude,
             y = temp_data$latitude,
             z = temp_data$logdur,
             main = toupper(state_list[i]))
}

```

We can see clumps of colors in both plots. This tells us we have evidence of spatial correalation in the data.

# Binned semicovariogram of log-duration for NY

#### Fields package

```{r echo = F, message = F, warning = F}

# Get data for ny
ny <- ufo %>%
  filter(state == 'ny')

# Create a semicovariogram for ny data
grd <- as.matrix(cbind(ny$longitude,ny$latitude))
v <- vgram(loc=grd,y=ny$logdur,N=100,lon.lat=TRUE)

# Create data frame of values
ny_v <- data.frame(Distance = v$d,
                   Value = sqrt(v$vgram))

# Add bins 
breaks <- seq(0,round(.98*max(v$d)/2),length.out = 100)
ny_v <- ny_v %>%
  mutate(bin = cut(Distance, breaks = breaks))

# Get statistics for summary        
ny_vstats <- data.frame(Centers = v$centers,
                        Mean = v$stats["mean",],
                        Med = v$stats["median",],
                        N = v$stats["N",])

# Filter out first half of esimates
ny_vstats <- ny_vstats %>%
  filter(Centers <= round(.98*max(v$d)/2))

```

```{r echo = F, message = F, warning = F}

# Create binned boxplot of the covariogram points
ggplot(data = ny_v, aes(x = as.factor(bin), y = Value)) +
  geom_boxplot() + 
  xlab('Distance Bin') +
  ylab('sqrt(Variogram)') + 
  scale_x_discrete(labels = round(breaks)) +
  ggtitle('Binned Covariogram Boxplot - 100 bins') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  stat_summary(fun.y=mean, geom="point", color = 'red')
```


```{r echo = F, message = F, warning = F}

# Plot the number of points per bin
ggplot(data = ny_vstats, aes(x = Centers, y = N)) +
  geom_point(color = 'blue') +
  xlab('Distance') +
  ylab('Count') +
  ggtitle('Number of points per bin') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0))
```

```{r echo = F, message = F, warning = F}

# Plot the variogram
ggplot(data = ny_vstats, aes(x = Centers, y = Mean)) +
  geom_point(color = 'blue') +
  xlab('Distance') +
  ylab('semivariance') +
  ggtitle('Binned semi-covariogram estimate') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  scale_x_continuous(limits = c(0, 150)) +
  geom_smooth(se = T, color = 'red')
```

There does not appear to be much spatial correlation in the data.

#### geoR package

```{r echo = F, message = F, warning = F}

# Technical difficulty: geoR only accepts projected coordinates
# Project coordinates
temp <- mapproject(x=grd[,1],y=grd[,2],projection="sinusoidal")
grd.p <- cbind(temp$x,temp$y)
rm(temp)

# Plot original vs. projected coordinates
par(mfrow=c(1,2))
plot(grd, main = 'Original', xlab = 'lon', ylab = 'lat')
plot(grd.p, main = 'Projected', xlab = 'proj1', ylab = 'proj2')

# Create breaks
breaks <- seq(0,max(rdist(grd.p)),length.out=500)

# Estimate the variogram classicly
v.c <- variog(coords = grd.p,
              data = ny$logdur, 
              estimator.type = "classical",
              op= "cloud") #cloud

# Estimate the variogram classicly via bins
v.bc <- variog(coords=grd.p,
               data=ny$logdur,
               estimator.type="classical",
               breaks=breaks,
               bin.cloud=TRUE) #binned

# Estimate the variogram robustsly
v.c.r <- variog(coords=grd.p,
                data=ny$logdur,
                estimator.type="modulus",
                op="cloud") #cloud

# Estimate the variogram robustsly via bins
v.bc.r <- variog(coords=grd.p,
                 data=ny$logdur,
                 estimator.type="modulus",
                 breaks=breaks,
                 bin.cloud=TRUE) #binned

# Compare classic vs. robust estimator
par(mfrow=c(1,2))
plot(v.bc,main="Classical-bin",pch=19,cex=0.5)
plot(v.bc.r,main="Robust-bin",pch=19,cex=0.5)

# We only car about the first half of the data
par(mfrow=c(1,2))
plot(v.bc, main="Classical-bin", pch=19, cex=0.5, 
     xlim = c(0,.08), ylim = c(3,5.5))
plot(v.bc.r, main="Robust-bin", pch=19, cex=0.5,
     xlim = c(0,.08), ylim = c(3,5.5))
```

#### Fitting a positive-definite functions

```{r echo = TRUE, message = F, warning = F}

# Fit several variograms
vfit1 <- variofit(v.bc.r,cov.model="exponential",weights="cressie")
vfit2 <- variofit(v.bc.r,cov.model="powered.exponential",weights="cressie",
                  nugget=0.3,fix.nugget=TRUE)
vfit3 <- variofit(v.bc.r,cov.model="matern",weights="cressie",kappa=1)
vfit4 <- variofit(v.bc.r,cov.model="spherical",weights="cressie")

# Plot the variogram
plot(v.bc.r)
lines(vfit1)
lines(vfit2,col="blue")
lines(vfit3,col="red")
lines(vfit4,col="green")

```

# Binned semicovariogram of log-duration for other state

#### Fields package

```{r echo = F, message = F, warning = F}

# Get data for ny
ny <- ufo %>%
  filter(state == 'ma')

# Create a semicovariogram for ny data
grd <- as.matrix(cbind(ny$longitude,ny$latitude))
v <- vgram(loc=grd,y=ny$logdur,N=100,lon.lat=TRUE)

# Create data frame of values
ny_v <- data.frame(Distance = v$d,
                   Value = sqrt(v$vgram))

# Add bins 
breaks <- seq(0,round(.98*max(v$d)/2),length.out = 100)
ny_v <- ny_v %>%
  mutate(bin = cut(Distance, breaks = breaks))

# Get statistics for summary        
ny_vstats <- data.frame(Centers = v$centers,
                        Mean = v$stats["mean",],
                        Med = v$stats["median",],
                        N = v$stats["N",])

# Filter out first half of esimates
ny_vstats <- ny_vstats %>%
  filter(Centers <= round(.98*max(v$d)/2))

```

```{r echo = F, message = F, warning = F}

# Create binned boxplot of the covariogram points
ggplot(data = ny_v, aes(x = as.factor(bin), y = Value)) +
  geom_boxplot() + 
  xlab('Distance Bin') +
  ylab('sqrt(Variogram)') + 
  scale_x_discrete(labels = round(breaks)) +
  ggtitle('Binned Covariogram Boxplot') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  stat_summary(fun.y=mean, geom="point", color = 'red')
```


```{r echo = F, message = F, warning = F}

# Plot the number of points per bin
ggplot(data = ny_vstats, aes(x = Centers, y = N)) +
  geom_point(color = 'blue') +
  xlab('Distance') +
  ylab('Count') +
  ggtitle('Number of points per bin') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0))
```

```{r echo = F, message = F, warning = F}

# Plot the variogram
ggplot(data = ny_vstats, aes(x = Centers, y = Mean)) +
  geom_point(color = 'blue') +
  xlab('Distance') +
  ylab('semivariance') +
  ggtitle('Binned semi-covariogram estimate') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  scale_x_continuous(limits = c(0, 150)) +
  geom_smooth(method = 'lm', se = F, color = 'red')
```

There does not appear to be much spatial correlation in the data.

#### geoR package

```{r echo = F, message = F, warning = F}

# Technical difficulty: geoR only accepts projected coordinates
# Project coordinates
temp <- mapproject(x=grd[,1],y=grd[,2],projection="sinusoidal")
grd.p <- cbind(temp$x,temp$y)
rm(temp)

# Plot original vs. projected coordinates
par(mfrow=c(1,2))
plot(grd, main = 'Original', xlab = 'lon', ylab = 'lat')
plot(grd.p, main = 'Projected', xlab = 'proj1', ylab = 'proj2')

# Create breaks
breaks <- seq(0,max(rdist(grd.p)),length.out=500)

# Estimate the variogram classicly
v.c <- variog(coords = grd.p,
              data = ny$logdur, 
              estimator.type = "classical",
              op= "cloud") #cloud

# Estimate the variogram classicly via bins
v.bc <- variog(coords=grd.p,
               data=ny$logdur,
               estimator.type="classical",
               breaks=breaks,
               bin.cloud=TRUE) #binned

# Estimate the variogram robustsly
v.c.r <- variog(coords=grd.p,
                data=ny$logdur,
                estimator.type="modulus",
                op="cloud") #cloud

# Estimate the variogram robustsly via bins
v.bc.r <- variog(coords=grd.p,
                 data=ny$logdur,
                 estimator.type="modulus",
                 breaks=breaks,
                 bin.cloud=TRUE) #binned

# Compare classic vs. robust estimator
par(mfrow=c(1,2))
plot(v.bc,main="Classical-bin",pch=19,cex=0.5)
plot(v.bc.r,main="Robust-bin",pch=19,cex=0.5)

# We only car about the first half of the data
par(mfrow=c(1,2))
plot(v.bc, main="Classical-bin", pch=19, cex=0.5, 
     xlim = c(0,.03), ylim = c(3,5))
plot(v.bc.r, main="Robust-bin", pch=19, cex=0.5,
     xlim = c(0,.03), ylim = c(3,5))
```

Not sure if this is a lack of sample conclusion

#### Fitting a positive-definite function

```{r echo = TRUE, message = F, warning = F}

# Fit several variograms
vfit <- variofit(v.bc.r,cov.model="exponential",weights="cressie")
vfit2 <- variofit(v.bc.r,cov.model="powered.exponential",weights="cressie",
  nugget=0.3,fix.nugget=TRUE)
vfit3 <- variofit(v.bc.r,cov.model="matern",weights="cressie",kappa=1)
vfit4 <- variofit(v.bc.r,cov.model="spherical",weights="cressie")

plot(v.bc.r, xlim = c(0,.03), ylim = c(3,5))
lines(vfit)
lines(vfit2,col="blue")
lines(vfit3,col="red")
lines(vfit4,col="green")
```

# Binned semicovariogram of log-duration for Arizona

#### Fields package

```{r echo = F, message = F, warning = F}

# Get data for ny
ny <- ufo %>%
  filter(state == 'az')

# Create a semicovariogram for ny data
grd <- as.matrix(cbind(ny$longitude,ny$latitude))
v <- vgram(loc=grd,y=ny$logdur,N=100,lon.lat=TRUE)

# Create data frame of values
ny_v <- data.frame(Distance = v$d,
                   Value = sqrt(v$vgram))

# Add bins 
breaks <- seq(0,round(.98*max(v$d)/2),length.out = 100)
ny_v <- ny_v %>%
  mutate(bin = cut(Distance, breaks = breaks))

# Get statistics for summary        
ny_vstats <- data.frame(Centers = v$centers,
                        Mean = v$stats["mean",],
                        Med = v$stats["median",],
                        N = v$stats["N",])

# Filter out first half of esimates
ny_vstats <- ny_vstats %>%
  filter(Centers <= round(.98*max(v$d)/2))

```

```{r echo = F, message = F, warning = F}

# Create binned boxplot of the covariogram points
ggplot(data = ny_v, aes(x = as.factor(bin), y = Value)) +
  geom_boxplot() + 
  xlab('Distance Bin') +
  ylab('sqrt(Variogram)') + 
  scale_x_discrete(labels = round(breaks)) +
  ggtitle('Binned Covariogram Boxplot') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  stat_summary(fun.y=mean, geom="point", color = 'red')
```


```{r echo = F, message = F, warning = F}

# Plot the number of points per bin
ggplot(data = ny_vstats, aes(x = Centers, y = N)) +
  geom_point(color = 'blue') +
  xlab('Distance') +
  ylab('Count') +
  ggtitle('Number of points per bin') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0))
```

```{r echo = F, message = F, warning = F}

# Plot the variogram
ggplot(data = ny_vstats, aes(x = Centers, y = Mean)) +
  geom_point(color = 'blue') +
  xlab('Distance') +
  ylab('semivariance') +
  ggtitle('Binned semi-covariogram estimate') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  scale_x_continuous(limits = c(0, 150)) +
  geom_smooth(method = 'lm', se = F, color = 'red')
```

There does not appear to be much spatial correlation in the data.

#### geoR package

```{r echo = F, message = F, warning = F}

# Technical difficulty: geoR only accepts projected coordinates
# Project coordinates
temp <- mapproject(x=grd[,1],y=grd[,2],projection="sinusoidal")
grd.p <- cbind(temp$x,temp$y)
rm(temp)

# Plot original vs. projected coordinates
par(mfrow=c(1,2))
plot(grd, main = 'Original', xlab = 'lon', ylab = 'lat')
plot(grd.p, main = 'Projected', xlab = 'proj1', ylab = 'proj2')

# Create breaks
breaks <- seq(0,max(rdist(grd.p)),length.out=500)

# Estimate the variogram classicly
v.c <- variog(coords = grd.p,
              data = ny$logdur, 
              estimator.type = "classical",
              op= "cloud") #cloud

# Estimate the variogram classicly via bins
v.bc <- variog(coords=grd.p,
               data=ny$logdur,
               estimator.type="classical",
               breaks=breaks,
               bin.cloud=TRUE) #binned

# Estimate the variogram robustsly
v.c.r <- variog(coords=grd.p,
                data=ny$logdur,
                estimator.type="modulus",
                op="cloud") #cloud

# Estimate the variogram robustsly via bins
v.bc.r <- variog(coords=grd.p,
                 data=ny$logdur,
                 estimator.type="modulus",
                 breaks=breaks,
                 bin.cloud=TRUE) #binned

# Compare classic vs. robust estimator
par(mfrow=c(1,2))
plot(v.bc,main="Classical-bin",pch=19,cex=0.5)
plot(v.bc.r,main="Robust-bin",pch=19,cex=0.5)

# We only car about the first half of the data
par(mfrow=c(1,2))
plot(v.bc, main="Classical-bin", pch=19, cex=0.5, 
     xlim = c(0,.04), ylim = c(3.5,5))
plot(v.bc.r, main="Robust-bin", pch=19, cex=0.5,
     xlim = c(0,.04), ylim = c(3,5.5))
```

#### Fitting a positive-definite functions

```{r echo = TRUE, message = F, warning = F}

# Fit several variograms
vfit1 <- variofit(v.bc.r,cov.model="exponential",weights="cressie")
vfit2 <- variofit(v.bc.r,cov.model="powered.exponential",weights="cressie",
                  nugget=0.3,fix.nugget=TRUE)
vfit3 <- variofit(v.bc.r,cov.model="matern",weights="cressie",kappa=1)
vfit4 <- variofit(v.bc.r,cov.model="spherical",weights="cressie")

# Plot the variogram
plot(v.bc.r, xlim = c(0,.04), ylim = c(3.5,5))
lines(vfit1)
lines(vfit2,col="blue")
lines(vfit3,col="red")
lines(vfit4,col="green")

```

# Strongest State Spatial Dependence

# Best Location for UFO Sighting

```{r echo = TRUE, message = F, warning = F}
```

```{r echo = TRUE, message = F, warning = F}
```